Climatic warming strengthens a positive feedback between alpine shrubs and fire

### 2016-03-14 manuscript/manuscript.Rnw
More edits

\begin{affiliations}
\item Department of Biological Sciences, Macquarie University, Sydney, NSW 2109, Australia.
\item The Centre of Excellence for Environmental Decisions, School of BioSciences, The University of Melbourne, Parkville 3010, Victoria, Australia.
\item Research Centre for Applied Alpine Ecology, La Trobe University, Melbourne 3086, Victoria, Australia.
\item CSIRO Land and Water, Tropical Ecosystems Research Centre, PMB 44 Winnellie 0821, NT, Australia.
\item Bio21 Institute, School of BioSciences, The University of Melbourne, Vic. 3010, Australia.
\textasteriskcentered Email for correspondence: \texttt{james.camac@gmail.com}
\end{affiliations}

@@

\subsection{Drivers of shrub seedling establishment.}
We first investigated how altitude, Topographic Wetness Index (TWI), adult density, fire and fire severity (as measured by minimum twig diameters\textemdash see Supplementary Methods) influenced the density of \textit{Grevillea} and \textit{Asterolasia} seedlings (the two dominant shrubs of alpine open heathland). In 2011-12, we estimated seedling density for both species across 40 alpine sites and found that the abundances of \textit{Grevillea} (Fig. \ref{fig2}) and \textit{Asterolasia} (Supplementary Fig. S1) seedlings were strongly influenced by the occurrence of fire. Sites burnt in 2003 had seedling densities between \Sexpr{density_ratios(asttry_seedling_density_model,'Asterolasia')} and \Sexpr{density_ratios(greaus_seedling_density_model,'Grevillea')} times higher than unburnt sites. The mean seedling density of \textit{Grevillea}, was \Sexpr{density_means(greaus_seedling_density_model,'Grevillea', TRUE)}/m$^2$ at burnt sites and \Sexpr{density_means(greaus_seedling_density_model,'Grevillea', FALSE)}/m$^2$ at unburnt sites. \textit{Asterolasia} had similar mean densities: \Sexpr{density_means(asttry_seedling_density_model,'Asterolasia', TRUE)} and \Sexpr{density_means(asttry_seedling_density_model, 'Asterolasia', FALSE)} seedlings/m$^2$ at burnt and unburnt sites, respectively. For both species, seedling density was also higher in severely burnt sites (i.e. sites with higher minimum twig diameters). As expected, pre-fire adult density positively influenced \textit{Grevillea} seedling density. For both shrub species, we detected no change in seedling density along a \Sexpr{round(max(complete_field_survey_data$altitude_m)) - round(min(complete_field_survey_data$altitude_m))} m altitudinal range (equivalent to a \Sexpr{round(0.77 * (200/100),1)}\textdegree C change in mean temperature\cite{Brown:1989vx}). The field warming experiment (see below) indicates that this lack of elevation response in \textit{Grevillea} is due to rates of recruitment (Supplementary Fig. S2) and mortality (see below) largely being insensitive to a \Sexpr{micro_clim(overall_microclimate_summary, 'CAmbient_Temp','mean','grand_mean')}\textdegree C change in temperature. We detected no significant effect of Topographic Wetness Index for either species.

@@

We used 40 long-term open heathland sites established after the 2003 fires\cite{Camac:2013eb} consisting of 30 burnt sites and 10 sites thought to be unburnt for over 70 years. In the summer of 2011-12, at each site, seedling density/m$^2$ was estimated using 40 quadrats, each 1 m$^2$, that were evenly distributed in groups of 10 along four 50 m transects, with 10 m between transect lines, subsampling an area of 2000 m$^2$. Within plots we recorded the number and maximum height of \textit{Grevillea} and \textit{Asterolasia} seedlings. For unburnt sites we counted the number of mature \textit{Grevillea} plants (\textgreater 0.5 m$^2$) within 5 m of each transect. In burnt sites, this required counting the number of skeletons (there were no living adults at any burnt site) that still persisted post-fire. We were unable to estimate numbers of adult \textit{Asterolasia} because this species does not have a persistent woody skeleton post-fire. Site level data, elevation and Topographic Wetness Index (TWI) were obtained from a 30 m resolution digital elevation model. Lastly, for burnt sites, fire severity was estimated by minimum twig diameters, collected immediately after the 2003 fires\cite{Camac:2013eb}.

@@

\begin{figure}
\begin{minipage}{\textwidth}
\includegraphics{figures/fig2.pdf}
\caption{\textbf{Mean \textit{Grevillea australis} seedling density along gradients of burning, fire severity (minimum twig diameter), altitude, Topographic Wetness Index (TWI) and adult density.} (\textbf{a}) Centered and standardized model coefficients (on the log scale); and effects of: (\textbf{b}) burning, (\textbf{c}) fire severity, (\textbf{d}) altitude, (\textbf{e}) topographic wetness and (\textbf{f}) adult density, in areas burnt by the 2003 fires. All bars and shaded areas indicate 95\% Bayesian Credible Intervals.}
\label{fig2}
\end{minipage}
\end{figure}

\begin{figure}
\begin{minipage}{\textwidth}
\includegraphics{figures/fig3.pdf}
\caption{\textbf{Observed and projected growth trajectories of four dominant Australian alpine shrubs.} (\textbf{a}) Observed height growth: Thick lines represent mean ($\pm$ 95\% confidence intervals) observed heights at each May census. Thin lines represent individual growth trajectories. (\textbf{b}) Mean ($\pm$ 95\% Bayesian Credible Interval) projected growth trajectories. In all cases, red \& orange lines = seedlings growing in warmed (OTC) conditions and blue and light blue lines = seedlings growing in control conditions. Projections were based on an logistic growth model using initial seedling size of 6 cm (the mean initial height observed in the OTC experiment) and mean maximum heights observed in long-unburnt sites (i.e. \Sexpr{hmax_values(hmax,'Asterolasia')}, \Sexpr{hmax_values(hmax,'Grevillea')}, \Sexpr{hmax_values(hmax,'Phebalium')} \& \Sexpr{hmax_values(hmax,'Prostanthera')} cm for \textit{Asterolasia}, \textit{Grevillea}, \textit{Phebalium}, \& \textit{Prostanthera}, respectively).}
\label{fig3}
\end{minipage}
\end{figure}

>>>>

\begin{affiliations}
\item Department of Biological Sciences, Macquarie University, Sydney, NSW 2109, Australia.
\item The Centre of Excellence for Environmental Decisions, School of BioSciences, The University of Melbourne, Melbourne, Victoria 3010, Australia.
\item Research Centre for Applied Alpine Ecology, La Trobe University, Melbourne, Victoria 3086, Australia.
\item CSIRO Land and Water, Tropical Ecosystems Research Centre, PMB 44 Winnellie, Northern Territory 0821, Australia.
\item Bio21 Institute, School of BioSciences, The University of Melbourne, Victoria 3010, Australia.
\textasteriskcentered Email for correspondence: \texttt{james.camac@gmail.com}
\end{affiliations}

@@

\subsection{Drivers of shrub seedling establishment.}
We first investigated how altitude, Topographic Wetness Index (TWI), adult density, fire and fire severity (as measured by twig diameters\textemdash see Supplementary Methods) influenced the density of \textit{Grevillea} and \textit{Asterolasia} seedlings (the two dominant shrubs of alpine open heathland). In 2011-12, we estimated seedling density for both species across 40 alpine sites and found that the abundances of \textit{Grevillea} (Fig. \ref{fig2}) and \textit{Asterolasia} (Supplementary Fig. S1) seedlings were strongly influenced by the occurrence of fire. Sites burnt in 2003 had seedling densities between \Sexpr{density_ratios(asttry_seedling_density_model,'Asterolasia')} and \Sexpr{density_ratios(greaus_seedling_density_model,'Grevillea')} times higher than unburnt sites. The mean seedling density of \textit{Grevillea}, was \Sexpr{density_means(greaus_seedling_density_model,'Grevillea', TRUE)}/m$^2$ at burnt sites and \Sexpr{density_means(greaus_seedling_density_model,'Grevillea', FALSE)}/m$^2$ at unburnt sites. \textit{Asterolasia} had similar mean densities: \Sexpr{density_means(asttry_seedling_density_model,'Asterolasia', TRUE)} and \Sexpr{density_means(asttry_seedling_density_model, 'Asterolasia', FALSE)} seedlings/m$^2$ at burnt and unburnt sites, respectively. For both species, seedling density was also higher in severely burnt sites (i.e. sites with higher twig diameters). As expected, pre-fire adult density positively influenced \textit{Grevillea} seedling density. For both shrub species, we detected no change in seedling density along a \Sexpr{round(max(complete_field_survey_data$altitude_m)) - round(min(complete_field_survey_data$altitude_m))} m altitudinal range (equivalent to a \Sexpr{round(0.77 * (200/100),1)}\textdegree C change in mean temperature\cite{Brown:1989vx}). The field warming experiment (see below) indicates that this lack of elevation response in \textit{Grevillea} is due to rates of recruitment (Supplementary Fig. S2) and mortality (see below) largely being insensitive to a \Sexpr{micro_clim(overall_microclimate_summary, 'CAmbient_Temp','mean','grand_mean')}\textdegree C change in temperature. We detected no significant effect of Topographic Wetness Index for either species.

@@

We used 40 long-term open heathland sites established after the 2003 fires\cite{Camac:2013eb} consisting of 30 burnt sites and 10 sites thought to be unburnt for over 70 years. In the summer of 2011-12, at each site, seedling density/m$^2$ was estimated using 40 quadrats, each 1 m$^2$, that were evenly distributed in groups of 10 along four 50 m transects, with 10 m between transect lines, subsampling an area of 2000 m$^2$. Within plots we recorded the number and maximum height of \textit{Grevillea} and \textit{Asterolasia} seedlings. For unburnt sites we counted the number of mature \textit{Grevillea} plants (\textgreater 0.5 m$^2$) within 5 m of each transect. In burnt sites, this required counting the number of skeletons (there were no living adults at any burnt site) that still persisted post-fire. We were unable to estimate numbers of adult \textit{Asterolasia} because this species does not have a persistent woody skeleton post-fire. Site level data, elevation and Topographic Wetness Index (TWI) were obtained from a 30 m resolution digital elevation model. Lastly, for burnt sites, fire severity was estimated by twig diameters, collected immediately after the 2003 fires\cite{Camac:2013eb}.

@@

\begin{figure}
\begin{minipage}{\textwidth}
\includegraphics{figures/fig2.pdf}
\caption{\textbf{Mean \textit{Grevillea australis} seedling density along gradients of burning, fire severity (twig diameter), altitude, Topographic Wetness Index (TWI) and adult density.} (\textbf{a}) Centered and standardized model coefficients (on the log scale); and effects of: (\textbf{b}) burning, (\textbf{c}) fire severity, (\textbf{d}) altitude, (\textbf{e}) topographic wetness and (\textbf{f}) adult density, in areas burnt by the 2003 fires. All bars and shaded areas indicate 95\% Bayesian Credible Intervals.}
\label{fig2}
\end{minipage}
\end{figure}

\begin{figure}
\begin{minipage}{\textwidth}
\includegraphics{figures/fig3.pdf}
\caption{\textbf{Observed and projected growth trajectories of four dominant Australian alpine shrubs.} (\textbf{a}) Observed height growth: Thick lines with error bars represent mean ($\pm$ 95\% confidence intervals) observed heights at each May census. Thin lines represent individual growth trajectories. (\textbf{b}) Mean ($\pm$ 95\% Bayesian Credible Interval) projected growth trajectories. In all cases, red and orange lines = seedlings growing in warmed (OTC) conditions and blue and light blue lines = seedlings growing in control conditions. Projections were based on an logistic growth model using initial seedling size of 6 cm (the mean initial height observed in the OTC experiment) and mean maximum heights observed in long-unburnt sites (i.e. \Sexpr{hmax_values(hmax,'Asterolasia')}, \Sexpr{hmax_values(hmax,'Grevillea')}, \Sexpr{hmax_values(hmax,'Phebalium')} \& \Sexpr{hmax_values(hmax,'Prostanthera')} cm for \textit{Asterolasia}, \textit{Grevillea}, \textit{Phebalium}, \& \textit{Prostanthera}, respectively).}
\label{fig3}
\end{minipage}
\end{figure}
